Accelerating Optimal Power Flow with GPUs: SIMD Abstraction of Nonlinear Programs and Condensed-Space Interior-Point Methods

### 2023-07-28 paper/main.tex
math typo

Now, we observe that a significant portion of the system in Fig.
\eqref{fig:very-long-eqn} can be eliminated by exploiting the block
structure, leading to an equivalent system stated in a smaller
space. In particular, the lower-right $4\times 4$ block is always
invertible since the IPM procedure ensures that the iterates stay in
the strict interior of the feasible set. This allows for eliminating
the lower-right 4x4 block, resulting in:
\begin{align}\label{eqn:very-long-reduced}
  &
    \underbrace{
    \begin{bmatrix}
      W  + \Sigma_x + \delta_w I & & A^{\top} \\
      & \Sigma_s + \delta_w I & -I\\
      A& -I & -\delta_c I\\
    \end{bmatrix}
  }_{M_{\text{aug}}}
  \begin{bmatrix}
    \Delta x \\
    \Delta s \\
    \Delta y \\
  \end{bmatrix}=
  \begin{bmatrix}
    q_x \\
    q_s\\
    q_y\\
  \end{bmatrix},
\end{align}
where:
\begin{align*}
  \Sigma_x&:= Z^\flat_x (X-X^\flat)^{-1}+ Z^\sharp_x (X^\sharp-X)^{-1}\\
  \Sigma_s&:= Z^\flat_s (S-S^\flat)^{-1}+ Z^\sharp_s (S^\sharp-S)^{-1}\\
  q_x&:=p_x + (X-X^\flat)^{-1} p_{z^\flat_x}-  (X^\sharp-X)^{-1} p_{z^\sharp_x}\\
  q_s&:= (S-S^\flat)^{-1} p_{z^\flat_s}-  (S^\sharp-S)^{-1} p_{z^\sharp_s}\\
  q_y&:=p_y.
\end{align*}
The bound dual steps can be recovered as follows:
\begin{align}\label{eqn:recover-1}
  \begin{aligned}[t]
    \Delta z^\flat_x &= \left(X-X^\flat\right)^{-1} \left(-Z^\flat_x \Delta x  + p_{z^\flat_x}\right)\\
    \Delta z^\sharp_x &= \left(X^\sharp-X\right)^{-1} \left(Z^\sharp_x \Delta x  + p_{z^\sharp_x}\right)\\
    \Delta z^\flat_s &= \left(S-S^\flat\right)^{-1} \left(-Z^\flat_s \Delta s  + p_{z^\flat_s}\right)\\
    \Delta z^\sharp_s &= \left(S^\sharp-S\right)^{-1} \left(Z^\sharp_s \Delta s  + p_{z^\sharp_s}\right).
  \end{aligned}
\end{align}
Note that the matrices involved in the inversions in \eqref{eqn:recover-1}
are always diagonal, so their computation is cheap.
Also, note that the augmented system in \eqref{eqn:very-long-reduced}
corresponds to the KKT system in \eqref{eqn:kkt-indefinite}. However,
in the original version of the algorithm \cite{pacaud2023accelerating}, we did not introduce the
slack variables, so it did not have the additional structure imposed
by the slack variables.

The key advantage of the inequality relaxation strategy is that it
imposes additional structure to the augmented KKT system, allowing us to
further reduce the dimension of the problem. In particular, the
lower-right 2x2 block in \eqref{eqn:very-long-reduced} can be
eliminated, which is a procedure called {\it condensation}; here, the invertibility of the lower-right block can be verified from the fact that $\delta_w,\delta_c\geq 0$ and $\Sigma_s\succ 0$. Through this,
we obtain the following system, written in the  primal
space only:
\begin{align}\label{eqn:very-long-condensed}
  (\underbrace{W + \delta_wI + \Sigma_x + A^{\top} D A}_{M_\text{cond}} ) \Delta x = q_x + A^\top (C q_y +  Dq_s ),
\end{align}
where:
\begin{align*}
  C := \left(\delta_c \Sigma_s + (1+\delta^{}_c\delta^{}_w) I\right)^{-1}, \;
  D := \left(\Sigma_s + \delta^{}_w I\right)C,
\end{align*}
and the dual and slack step directions can be recovered by:
\begin{align}
  \Delta s &:= C \left(\delta_c q_s - (q_y + A\Delta x)\right)\nonumber\\
  \Delta y &:= (\Sigma_s + \delta_w I) \Delta s -q_s.\label{eqn:recover-2}
\end{align}
Again, the matrices involved in the inversions above are always
diagonal, so their computation is cheap.

>>>>

Now, we observe that a significant portion of the system in Fig.
\eqref{fig:very-long-eqn} can be eliminated by exploiting the block
structure, leading to an equivalent system stated in a smaller
space. In particular, the lower-right $4\times 4$ block is always
invertible since the IPM procedure ensures that the iterates stay in
the strict interior of the feasible set. This allows for eliminating
the lower-right 4x4 block, resulting in:
\begin{align}\label{eqn:very-long-reduced}
  &
    \underbrace{
    \begin{bmatrix}
      W  + \Sigma_x + \delta_w I & & A^{\top} \\
      & \Sigma_s + \delta_w I & -I\\
      A& -I & -\delta_c I\\
    \end{bmatrix}
  }_{M_{\text{aug}}}
  \begin{bmatrix}
    \Delta x \\
    \Delta s \\
    \Delta y \\
  \end{bmatrix}=
  \begin{bmatrix}
    q_x \\
    q_s\\
    q_y\\
  \end{bmatrix},
\end{align}
where:
\begin{align*}
  \Sigma_x&:= Z^\flat_x (X-X^\flat)^{-1}+ Z^\sharp_x (X^\sharp-X)^{-1}\\
  \Sigma_s&:= Z^\flat_s (S-S^\flat)^{-1}+ Z^\sharp_s (S^\sharp-S)^{-1}\\
  q_x&:=p_x + (X-X^\flat)^{-1} p_{z^\flat_x}-  (X^\sharp-X)^{-1} p_{z^\sharp_x}\\
  q_s&:=p_s + (S-S^\flat)^{-1} p_{z^\flat_s}-  (S^\sharp-S)^{-1} p_{z^\sharp_s}\\
  q_y&:=p_y.
\end{align*}
The bound dual steps can be recovered as follows:
\begin{align}\label{eqn:recover-1}
  \begin{aligned}[t]
    \Delta z^\flat_x &= \left(X-X^\flat\right)^{-1} \left(-Z^\flat_x \Delta x  + p_{z^\flat_x}\right)\\
    \Delta z^\sharp_x &= \left(X^\sharp-X\right)^{-1} \left(Z^\sharp_x \Delta x  + p_{z^\sharp_x}\right)\\
    \Delta z^\flat_s &= \left(S-S^\flat\right)^{-1} \left(-Z^\flat_s \Delta s  + p_{z^\flat_s}\right)\\
    \Delta z^\sharp_s &= \left(S^\sharp-S\right)^{-1} \left(Z^\sharp_s \Delta s  + p_{z^\sharp_s}\right).
  \end{aligned}
\end{align}
Note that the matrices involved in the inversions in \eqref{eqn:recover-1}
are always diagonal, so their computation is cheap.
Also, note that the augmented system in \eqref{eqn:very-long-reduced}
corresponds to the KKT system in \eqref{eqn:kkt-indefinite}. However,
in the original version of the algorithm \cite{pacaud2023accelerating}, we did not introduce the
slack variables, so it did not have the additional structure imposed
by the slack variables.

The key advantage of the inequality relaxation strategy is that it
imposes additional structure to the augmented KKT system, allowing us to
further reduce the dimension of the problem. In particular, the
lower-right 2x2 block in \eqref{eqn:very-long-reduced} can be
eliminated, which is a procedure called {\it condensation}; here, the invertibility of the lower-right block can be verified from the fact that $\delta_w,\delta_c\geq 0$ and $\Sigma_s\succ 0$. Through this,
we obtain the following system, written in the  primal
space only:
\begin{align}\label{eqn:very-long-condensed}
  (\underbrace{W + \delta_wI + \Sigma_x + A^{\top} D A}_{M_\text{cond}} ) \Delta x = q_x + A^\top (C q_s +  Dq_y ),
\end{align}
where:
\begin{align*}
  C := \left(\delta_c \Sigma_s + (1+\delta^{}_c\delta^{}_w) I\right)^{-1}, \;
  D := \left(\Sigma_s + \delta^{}_w I\right)C,
\end{align*}
and the dual and slack step directions can be recovered by:
\begin{align}
  \Delta s &:= C \left(\delta_c q_s - (q_y + A\Delta x)\right)\nonumber\\
  \Delta y &:= (\Sigma_s + \delta_w I) \Delta s -q_s.\label{eqn:recover-2}
\end{align}
Again, the matrices involved in the inversions above are always
diagonal, so their computation is cheap.
